Effectful Semantics for Substructural SSA

### 2023-11-14 pldi24/paper.tex
Tap

@@

\begin{definition}[Binoidal Category] 
  A \textbf{binoidal category} \(\mc{C}\) is a category equipped with a binary
  operation on objects \(\otimes: \mc{C} \times \mc{C} \to \mc{C}\) which is
  independently functorial in each argument (i.e., for each object \(A\), \(A
  \otimes -\) and \(- \otimes A\) are functors).
\end{definition}
Given \(f: \mc{C}(A, B)\), \(g: \mc{C}(A', B')\) in a binoidal category, we
write \(f \ltimes g = f \otimes A';B \otimes g\) and \(f \rtimes g = A \otimes
g;f \otimes B'\); we say \(f\) is \textbf{central} if, for all \(g\), \(f
\ltimes g = f \rtimes g\) and \(g \ltimes f = g \rtimes f\), in which case we
simply write \(f \otimes g\), \(g \otimes f\) respectively.

\begin{definition}[Premonoidal Category] 
  A \textbf{premonoidal category} is a binoidal category with an identity object
  \(I: \mc{C}\) such that
  \begin{itemize}
    \item There exist central natural isomorphisms \(\lambda_A: I \otimes A
    \simeq A\) (the \textbf{left unitor}), \(\rho_A: A \otimes I \simeq A\) (the
    \textbf{right unitor})
    \item There exists a central isomorphism \(\alpha_{A, B, C}: (A \otimes B)
    \otimes C \simeq A \otimes (B \otimes C)\) (the \textbf{associator})
    \item Satisfying the \textbf{pentagon law}: \todo{this}
    \item Satisfying the \textbf{triangle law}: \todo{this}
  \end{itemize}
  We say a premonoidal category is \textbf{symmetric} if it is equipped with a
  central natural isomorphism \(\sigma_{A, B}: A \otimes B \simeq B \otimes A\)
  (the \textbf{symmetry}) satisfying the \textbf{hexagon law} \todo{this}.

  We say a premonoidal category is \textbf{monoidal} if every morphism is
  central.
\end{definition}

\begin{definition}[Effectful Category] \
  \TODO{this}
\end{definition}

@@

% \TODO{text}

@@

\TODO{Hyland and Hasegawa traces?}

\TODO{intuition? Can draw diagrams, but those take up space}

>>>>

In this section, we will give a categorical semantics for \isotopessa. Starting
with Freyd categories, we will generalize to \textit{substructural effectful
categories} to support substructurality. We will then require our substructural
effectful category to have \textit{distributive coproducts} to implement
branching. Finally, we will require an \textit{Elgot structure} to implement
general, potentially nonterminating control flow.

@@

We begin with some basic definitions:
\begin{definition}[Binoidal Category] 
  A \textbf{binoidal category} \(\mc{C}\) is a category equipped with a binary
  operation on objects \(\otimes: \mc{C} \times \mc{C} \to \mc{C}\) which is
  independently functorial in each argument (i.e., for each object \(A\), \(A
  \otimes -\) and \(- \otimes A\) are functors).
\end{definition}
Given \(f: \mc{C}(A, B)\), \(g: \mc{C}(A', B')\) in a binoidal category, we
write \(f \ltimes g = f \otimes A';B \otimes g\) and \(f \rtimes g = A \otimes
g;f \otimes B'\); we say \(f\) is \textbf{central} if, for all \(g\), \(f
\ltimes g = f \rtimes g\) and \(g \ltimes f = g \rtimes f\), in which case we
simply write \(f \otimes g\), \(g \otimes f\) respectively.
\begin{definition}[Premonoidal Category] 
  A \textbf{premonoidal category} is a binoidal category with an identity object
  \(I: \mc{C}\) such that
  \begin{itemize}
    \item There exist central natural isomorphisms \(\lambda_A: I \otimes A
    \simeq A\) (the \textbf{left unitor}), \(\rho_A: A \otimes I \simeq A\) (the
    \textbf{right unitor})
    \item There exists a central isomorphism \(\alpha_{A, B, C}: (A \otimes B)
    \otimes C \simeq A \otimes (B \otimes C)\) (the \textbf{associator})
    \item Satisfying the \textbf{pentagon law}: \todo{this}
    \item Satisfying the \textbf{triangle law}: \todo{this}
  \end{itemize}
  We say a premonoidal category is \textbf{symmetric} if it is equipped with a
  central natural isomorphism \(\sigma_{A, B}: A \otimes B \simeq B \otimes A\)
  (the \textbf{symmetry}) satisfying the \textbf{hexagon law} \todo{this}. We
  say a premonoidal category is \textbf{monoidal} if every morphism is central.
\end{definition}
One issue with the notion of a premonoidal category is that it is unclear what
exactly a premonoidal functor should be: if we want them to compose, we would
like to preserve central morphisms, but trying to preserve \textit{all} central
morphisms is too strict a notion. We want, in essence, to be able to distinguish
morphisms that are truly "pure" from those which simply happen to be central. To
do that, we introduce the notion of an \textit{effectful category} as follows:
\begin{definition}[Effectful Category] 
  An \textbf{effectful category} \(\mc{V} \to \mc{C}\) is an identity on objects
  functor \(\upg{\cdot}{}: \mc{V} \to \mc{C}\) from a monoidal category
  \(\mc{V}\) to a premonoidal category \(\mc{C}\) which preserves all
  premonoidal structure (i.e. sends associators to associators, unitors to
  unitors, and tensor products to tensor products). It is \textbf{symmetric} if
  both \(\mc{V}\) and \(\mc{C}\) are, and \(\upg{\cdot}{}\) preserves the
  symmetry. A \textbf{Freyd category} is an effectful category where \(\mc{V}\)
  is Cartesian.
\end{definition}

\TODO{text on Freyd categories}

@@

@@
